An Experimental Evaluation of Hybrid Querying on Vectors (EA\&B)

### 2025-05-01 VLDB2026/main.tex
Update main.tex

	In recent years, hybrid query algorithms develop rapidly, giving rise to numerous attribute filtering ~\cite{NHQ,diskann} and range filtering algorithms \cite{serf,iRangeGraph}. In practical applications, the performance of hybrid query algorithms is influenced not only by unstructured data but also closely related to structured data. For attribute filtering algorithms, factors such as \textcolor{blue}{the number of attributes}, attribute distribution \cite{UNG}, and attribute selectivity \cite{ACORN} significantly impact algorithm performance. As for range filtering algorithms, the size of the query range and the characteristics of different datasets  play an important role in determining algorithm performance.

	Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some ANN algorithms, it presents several notable limitations:
	1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. 
	\sout{For instance, attribute filtering algorithms such as NHQ and UNG are not included in the evaluation. }
	2) It \textcolor{blue}{focuses only on single-attribute and dual-attribute query scenarios,} 
	\sout{neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes.}\textcolor{blue}{ \sout{For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.}}
	3) It does not evaluate algorithm performance under different attribute distributions. 
	\sout{However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.}

@@

	(1)\textbf{ Systematic Classification and Overview.}
	%We systematically classify 6 representative attribute filtering algorithms from multiple dimensions, including index organization, filtering strategy, Boolean logic support, and index construction methods. Additionally, we survey 5 range filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts provide a unified reference framework for future study.
	\textcolor{red}{We systematically classify 6 representative attribute filtering algorithms along multiple dimensions, including index organization, filtering strategies, Boolean logic support, and index construction methods. In addition, we survey 5 range filtering algorithms, and provide an overview of widely used 4 vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts offer a unified reference framework for future research.}

	\textcolor{blue}{
		(2) \textbf{Enhancing Datasets and Experimental Settings for Fair Evaluation.}
		To address the lack of standardized benchmarks in attribute filtering research, we enrich commonly used datasets by generating attribute values tailored to real-world scenarios. Furthermore, we design comprehensive experimental settings that reflect diverse application requirements, including varying attribute distributions, selectivity, and query conditions. These enhancements provide a unified evaluation framework that supports fair, consistent, and reproducible comparisons across different algorithms, laying a solid foundation for future research in hybrid query.
	}

	(3)\textbf{ Evaluation of Attribute Filtering Algorithms.}
	%We conduct a systematic evaluation of 13 attribute filtering algorithms on 7 real-world datasets. 
	\textcolor{red}{We systematically evaluate 6 attribute filtering algorithms, 1 library, and 3 databases on 7 real-world datasets.} By analyzing performance under varying numbers of attributes, we reveal the strengths and weaknesses of each method. We further examine their behavior under different attribute distributions and selectivity to assess robustness and adaptability in complex query scenarios. Evaluation metrics include index construction time, index size, peak memory usage, QPS, and search accuracy.

	(4)\textbf{ Evaluation of Range Filtering Algorithms.}
	%We benchmark 5 range filtering algorithms\textcolor{red}{, 1 library, and 3 databases} 
	\textcolor{red}{We benchmark algorithms that support range filtering} on 3 large-scale datasets, using varying query range settings in the experiments. The experimental results \textcolor{blue}{show} the performance of these algorithms in terms of index construction efficiency, storage overhead, and query performance. \textcolor{blue}{We further analyze the factors that affect the performance of the range query algorithm.}

	(5)\textbf{ Recommendations and Challenges.}
	Based on the experimental results, we provide algorithm selection recommendations for common application scenarios and highlight key challenges in the field of hybrid queries. These challenges include limited Boolean logic support, the lack of multi-attribute range filtering capabilities, and the sensitivity of algorithms to data distribution. Currently, few methods simultaneously support both attribute filtering and range filtering, pointing to potential directions for future research.

@@

	Current hybrid query methods mainly adopt graph-based ~\cite{nsw,kgraph,nsg,fanng,ngt} or Inverted File Index (IVF)-based \cite{PQ} index organization. We briefly introduce the fundamental principles of these two types in the following.

@@

	\section{Hybrid query Algorithms}

@@

	\noindent\textbf{\underline{ACORN \cite{ACORN}.}}  
	%	Above simultaneous filtering algorithms struggle with large-scale, unbounded, or unknown predicate sets.
	%	%The above simultaneous filtering methods are less effective in scenarios involving large-scale, unbounded, or unknown predicate sets, where a predicate refers to a condition or rule used for filtering. 
	%	ACORN addresses this issue by introducing a predicate-agnostic indexing framework.
	%	ACORN is built upon Hierarchical Navigable Small World (HNSW) \cite{hnsw}. It supports high-cardinality and unrestricted predicate sets, overcoming the limitations of methods restricted to small-scale equality filters. It constructs a denser hierarchical graph by expanding node neighborhoods and prunes edges at lower levels to control index size. During the query process, ACORN eliminates unnecessary distance computations by filtering out nodes that violate attribute constraints. This effectively maintains a nearest neighbor graph over valid nodes only. 
	%	%During the query process, ACORN avoids unnecessary distance computations by filtering out nodes that do not meet attribute constraints, effectively simulating a nearest neighbor graph over valid nodes only.
	%	ACORN includes 2 indexes: 1) \textbf{ACORN-$\gamma$}, which expands neighbor lists during construction, increasing memory usage for higher performance.  
	%	2) \textbf{ACORN-1}, which extends neighbor lists by considering second-hop neighbors during search, reducing index size with minor performance loss.
	%Above simultaneous filtering algorithms struggle with large-scale, unbounded, or unknown predicate sets. ACORN addresses this by introducing a predicate-agnostic indexing framework.
	\textcolor{red}{Pre-filtering methods apply attribute constraints prior to ANN search. As a pre-filtering approach, ACORN adopts a predicate-agnostic indexing framework to overcome the limitations of the aforementioned simultaneous filtering methods in handling large-scale or unknown predicate settings.}
	%Built upon Hierarchical Navigable Small World (HNSW) \cite{hnsw}, ACORN supports high-cardinality and unrestricted predicates.
	%, overcoming limitations of methods confined to small-scale equality filters. 
	%It constructs a denser hierarchical graph by expanding neighborhoods and prunes lower-level edges to control index size. 
	%During querying, ACORN eliminates unnecessary distance computations by filtering out nodes violating attribute constraints, effectively maintaining a nearest neighbor graph over valid nodes only. 
	\textcolor{red}{Based on Hierarchical Navigable Small World (HNSW) \cite{hnsw}, ACORN constructs a denser hierarchical graph with neighborhood expansion and edge pruning. During search, it filters out nodes violating constraints, maintaining a valid nearest neighbor graph.}
	ACORN includes two indexes: 1) \textbf{ACORN-$\gamma$}, which expands neighbor lists during construction, trading memory for higher performance; 
	2) \textbf{ACORN-1}, which extends neighbor lists via second-hop neighbors during search, reducing index size with minor performance loss.

@@

	\noindent\textbf{\underline{SeRF \cite{serf}.}} Conventional RF-ANN search approaches typically follow one of two paradigms: 1) Conducting ANN search first followed by attribute-based filtering. 2) Filtering the dataset based on attribute ranges before performing ANN search. Building a separate neighbor graph (e.g., HNSW) for each attribute range could ensure efficient querying but incurs an $O(n^2)$ cost in constructing and storing $n$ graphs. Since many edges are shared across these graphs, SeRF introduces a validity-range aware design where each edge records the interval in which it is valid, indicating in which subgraphs the edge remains valid. This compresses $n$ graphs into a single unified graph, maintaining search effectiveness while significantly reducing memory and index construction overhead.

@@

	\noindent\textbf{\underline{UNIFY \cite{UNIFY}.}}  
	 Unlike DSG, which relies on rectangle trees and range-aware edges, UNIFY adopts a segmentation-based approach and proposes a range query index structure. UNIFY introduces the Segmented Inclusive Graph (SIG), which partitions the dataset into segments by attribute and constructs an independent neighbor graph for each segment. These segment graphs are then integrated into a unified global graph. Its hierarchical variant, HSIG, enhances SIG by incorporating the multilayer design of HNSW, skip lists, and edge bitmaps for efficient range localization and post-filtering pruning. UNIFY supports 3 filtering strategies—pre-filtering, post-filtering, and  simultaneous filtering—enabling robust and flexible performance.

@@

	\subsubsection{Robustness}In the following, we evaluate the robustness of these algorithms.

	\textit{\textbf{Attribute Distribution.}} \textcolor{blue}{To evaluate} performance across different attribute distributions, we generate base and query attribute sets with 4 representative distributions—long-tailed, normal, power-law, and uniform—across the 6 datasets described in Section~4.1.1. We evaluate all algorithms under these distributions on each dataset. The performance differences observed for the same algorithm across different datasets are relatively minor. Thus, due to space constraints, we present only the results on SIFT1M.

>>>>

	In recent years, hybrid query algorithms develop rapidly, giving rise to numerous attribute filtering ~\cite{NHQ,diskann} and range filtering algorithms \cite{serf,iRangeGraph}. In practical applications, the performance of hybrid query algorithms is influenced not only by unstructured data but also closely related to structured data. For attribute filtering algorithms, factors such as the number of base attributes, the number of query attributes, attribute distribution \cite{UNG}, and attribute selectivity \cite{ACORN} significantly impact algorithm performance. As for range filtering algorithms, the size of the query range and the characteristics of different datasets also play an important role in determining algorithm performance.

	Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some algorithms, it presents several notable limitations:
	1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. 
	\sout{For instance, attribute filtering algorithms such as NHQ and UNG are not included in the evaluation. }
	2) It \textcolor{blue}{focuses only on single-attribute and dual-attribute query scenarios,} 
	\sout{neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes.}\textcolor{blue}{ \sout{For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.}}
	3) It does not evaluate algorithm performance under different attribute distributions. 
	\sout{However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.}

@@

	(1)\textbf{ Systematic Classification and Overview.}
	%We systematically classify 6 representative attribute filtering algorithms from multiple dimensions, including index organization, filtering strategy, Boolean logic support, and index construction methods. Additionally, we survey 5 range filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts provide a unified reference framework for future study.
	\textcolor{red}{We systematically classify 6 representative attribute filtering algorithms along multiple dimensions, including index organization, filtering strategies, Boolean logic support, and index construction methods. In addition, we survey 5 range filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts offer a unified reference framework for future research.}

	\textcolor{blue}{
		(2) \textbf{Enhancing Datasets and Experimental Settings for Fair Evaluation.}
		To address the lack of standardized benchmarks in attribute filtering research, we enrich commonly used datasets by generating attribute values tailored to real-world scenarios. Furthermore, we design comprehensive experimental settings that reflect diverse application requirements, including varying attribute distributions, selectivity levels, and query conditions. These enhancements provide a unified evaluation framework that supports fair, consistent, and reproducible comparisons across different algorithms, laying a solid foundation for future research in hybrid query processing.
	}

	(3)\textbf{ Evaluation of Attribute Filtering Algorithms.}
	%We conduct a systematic evaluation of 13 attribute filtering algorithms on 7 real-world datasets. 
	\textcolor{red}{We systematically evaluate 6 attribute filtering algorithms, 1 library, and 3 databases on 7 real-world datasets.}By analyzing performance under varying numbers of attributes, we reveal the strengths and weaknesses of each method. We further examine their behavior under different attribute distributions and selectivity conditions to assess robustness and adaptability in complex query scenarios. Evaluation metrics include index construction time, index size, peak memory usage, QPS, and search accuracy.

	(4)\textbf{ Evaluation of Range Filtering Algorithms.}
	%We benchmark 5 range filtering algorithms\textcolor{red}{, 1 library, and 3 databases} 
	\textcolor{red}{We benchmark 5 algorithms specifically designed for range filtering, along with 5 others applicable to range filtering} on 3 large-scale datasets, using varying query range settings in the experiments. The experimental results demonstrate the performance of these algorithms in terms of index construction efficiency, storage overhead, and query performance. We further analyze performance factors from an algorithmic perspective.

	(5)\textbf{ Recommendations and Challenges.}
	Based on the experimental results, we provide algorithm selection recommendations for common application scenarios and highlight key challenges in the field of hybrid queries. These challenges include limited Boolean logic support, the lack of multi-attribute range filtering capabilities, and the sensitivity of algorithms to data distribution. \sout {Currently, few methods simultaneously support both attribute filtering and range filtering, pointing to potential directions for future research.}

@@

	Current hybrid query methods mainly adopt graph-based ~\cite{nsw,kgraph,nsg,fanng,ngt} or Inverted File Index (IVF)-based \cite{PQ} index organization. We briefly introduce the fundamental principles of these 2 types in the following.

@@

	\section{Overview of Hybrid query Algorithms}

@@

	\noindent\textbf{\underline{ACORN \cite{ACORN}.}}  
	%	Above simultaneous filtering algorithms struggle with large-scale, unbounded, or unknown predicate sets.
	%	%The above simultaneous filtering methods are less effective in scenarios involving large-scale, unbounded, or unknown predicate sets, where a predicate refers to a condition or rule used for filtering. 
	%	ACORN addresses this issue by introducing a predicate-agnostic indexing framework.
	%	ACORN is built upon Hierarchical Navigable Small World (HNSW) \cite{hnsw}. It supports high-cardinality and unrestricted predicate sets, overcoming the limitations of methods restricted to small-scale equality filters. It constructs a denser hierarchical graph by expanding node neighborhoods and prunes edges at lower levels to control index size. During the query process, ACORN eliminates unnecessary distance computations by filtering out nodes that violate attribute constraints. This effectively maintains a nearest neighbor graph over valid nodes only. 
	%	%During the query process, ACORN avoids unnecessary distance computations by filtering out nodes that do not meet attribute constraints, effectively simulating a nearest neighbor graph over valid nodes only.
	%	ACORN includes 2 indexes: 1) \textbf{ACORN-$\gamma$}, which expands neighbor lists during construction, increasing memory usage for higher performance.  
	%	2) \textbf{ACORN-1}, which extends neighbor lists by considering second-hop neighbors during search, reducing index size with minor performance loss.
	%Above simultaneous filtering algorithms struggle with large-scale, unbounded, or unknown predicate sets. ACORN addresses this by introducing a predicate-agnostic indexing framework.
	\textcolor{red}{Pre-filtering methods apply attribute constraints prior to ANN search. As a pre-filtering approach, ACORN adopts a predicate-agnostic indexing framework to overcome the limitations of the aforementioned simultaneous filtering methods in handling large-scale or unknown predicate settings. }
	%Built upon Hierarchical Navigable Small World (HNSW) \cite{hnsw}, ACORN supports high-cardinality and unrestricted predicates.
	%, overcoming limitations of methods confined to small-scale equality filters. 
	%It constructs a denser hierarchical graph by expanding neighborhoods and prunes lower-level edges to control index size. 
	%During querying, ACORN eliminates unnecessary distance computations by filtering out nodes violating attribute constraints, effectively maintaining a nearest neighbor graph over valid nodes only. 
		\textcolor{red}{Based on Hierarchical Navigable Small World (HNSW) \cite{hnsw}, ACORN constructs a denser hierarchical graph with neighborhood expansion and edge pruning.During search, it filters out nodes violating constraints, maintaining a valid nearest neighbor graph.}
	ACORN includes two indexes: 1) \textbf{ACORN-$\gamma$}, which expands neighbor lists during construction, trading memory for higher performance; 
	2) \textbf{ACORN-1}, which extends neighbor lists via second-hop neighbors during search, reducing index size with minor performance loss.

@@

	\noindent\textbf{\underline{SeRF \cite{serf}.}}  
	Conventional RF-ANN search approaches typically follow one of two paradigms: 1) Conducting ANN search first followed by attribute-based filtering. 2) Filtering the dataset based on attribute ranges before performing ANN search. \sout{However, both approaches suffer from suboptimal performance.} Building a separate neighbor graph (e.g., HNSW) for each attribute range could ensure efficient querying but incurs an $O(n^2)$ cost in constructing and storing $n$ graphs. Since many edges are shared across these graphs, SeRF introduces a validity-range aware design where each edge records the interval in which it is valid, indicating in which subgraphs the edge remains valid. This compresses $n$ graphs into a single unified graph, maintaining search effectiveness while significantly reducing memory and index construction overhead.

@@

	\noindent\textbf{\underline{UNIFY \cite{UNIFY}.}}  
	Unlike DSG, which relies on rectangle trees and range-aware edges, UNIFY adopts a segmentation-based approach and proposes a range query index structure. UNIFY introduces the Segmented Inclusive Graph (SIG), which partitions the dataset into segments by attribute and constructs an independent neighbor graph for each segment. These segment graphs are then integrated into a unified global graph. \sout{SIG adheres to a graph inclusiveness principle: any query subgraph can be composed from existing segment graphs, avoiding the need for range-specific index construction.} Its hierarchical variant, HSIG, enhances SIG by incorporating the multilayer design of HNSW, skip lists, and edge bitmaps for efficient range localization and post-filtering pruning. UNIFY supports 3 filtering strategies—pre-filtering, post-filtering, and  simultaneous filtering—enabling robust and flexible performance.

@@

	\subsubsection{Robustness}In the following, we evaluate the robustness of \textcolor{blue}{attribute filtering algorithms.}

	\textit{\textbf{Attribute Distribution.}} \textcolor{blue}{To evaluate the performance of these algorithms across different attribute distributions, we generate attribute} sets with 4 representative distributions—long-tailed, normal, power-law, and uniform—across the 6 datasets described in Section~4.1.1. We evaluate all algorithms under these distributions on each dataset. The performance differences observed for the same algorithm across different datasets are relatively minor. Thus, due to space constraints, we present only the results on SIFT1M.
